\begin{thebibliography}{10}

\bibitem{benanav87jsc}
D.~Benanav, D.~Kapur, and P.~Narendran.
\newblock Complexity of matching problems.
\newblock {\em J. of Symbolic Computation}, 3(1-2):203--216, 1987.

\bibitem{bendix70book}
P.~Bendix and D.~Knuth.
\newblock {\em Computational problems in abstract algebra}, chapter Simple word
  problems in universal algebra.
\newblock Pergamon Press, 1970.

\bibitem{burton93jfp}
F.~Burton and R.~Cameron.
\newblock Pattern matching with abstract data types.
\newblock {\em J. of Functional Programming}, 3(2):171--190, 1993.

\bibitem{burton96}
W.~Burton, E.~Meijer, P.~Sansom, S.~Thompson, and P.~Wadler.
\newblock Views: An extension to {H}askell pattern matching.
\newblock \url{http://www.haskell.org/extensions/views.html}, 1996.

\bibitem{lechenadec86book}
P.~Le Chenadec.
\newblock {\em Canonical forms in finitely presented algebras}.
\newblock Research notes in theoretical computer science. Pitman, 1986.

\bibitem{cime2.02}
E.~Contejean, C.~March\'e, B.~Monate, and X.~Urbain.
\newblock {\em {CiME} version 2.02}.
\newblock LRI, CNRS UMR 8623, Université Paris-Sud, France, 2004.
\newblock \url{http://cime.lri.fr/}.

\bibitem{coq8.0}
{Coq Development Team}.
\newblock {\em The Coq Proof Assistant Reference Manual, Version 8.0}.
\newblock INRIA, France, 2006.
\newblock \url{http://coq.inria.fr/}.

\bibitem{dershowitz90book}
N.~Dershowitz and J.-P. Jouannaud.
\newblock Rewrite systems.
\newblock In J.~van Leeuwen, editor, {\em Handbook of Theoretical Computer
  Science}, volume~B, chapter~6. North Holland, 1990.

\bibitem{zenon0.4.1}
D.~Doligez.
\newblock Zenon, version 0.4.1.
\newblock \url{http://focal.inria.fr/zenon/}, 2006.

\bibitem{ocaml3.09}
D.~Doligez, J.~Garrigue, X.~Leroy, D.~Rémy, and J.~Vouillon.
\newblock {\em The Objective Caml system release 3.09, Documentation and user's
  manual}.
\newblock INRIA, France, 2005.
\newblock \url{http://caml.inria.fr/}.

\bibitem{haskell03}
S.~P.~Jones (editor).
\newblock {\em Haskell 98 Language and Libraries, The revised report}.
\newblock Cambridge University Press, 2003.

\bibitem{gaillourdet03cade}
J.-M. Gaillourdet, T.~Hillenbrand, B.~Löchner, and H.~Spies.
\newblock The new {W}aldmeister loop at work.
\newblock In {\em Proc. of CADE'03\em, LNCS 2741}.
\newblock \url{http://www.waldmeister.org/}.

\bibitem{hullot80thesis}
J.-M. Hullot.
\newblock {\em Compilation de formes canoniques dans les théories
  équationnelles}.
\newblock PhD thesis, Universit\'e Paris 11, France, 1980.

\bibitem{jouannaud06rta}
J.-P. Jouannaud.
\newblock Modular church-rosser modulo.
\newblock In {\em Proc. of RTA'06\em, LNCS 4098}.

\bibitem{tom2.3}
P.-E. Moreau, E.~Balland, P.~Brauner, R.~Kopetz, and A.~Reilles.
\newblock {\em Tom Manual version 2.3}.
\newblock INRIA \& LORIA, Nancy, France, 2006.
\newblock \url{http://tom.loria.fr/}.

\bibitem{moreau03cc}
P.-E. Moreau, C.~Ringeissen, and M.~Vittek.
\newblock A pattern matching compiler for multiple target languages.
\newblock In {\em Proc. of CC'03\em, LNCS 2622}.

\bibitem{okasaki98ml}
C.~Okasaki.
\newblock Views for standard {ML}.
\newblock In {\em Proc. of ML'98}.

\bibitem{peterson81jacm}
G.~Peterson and M.~Stickel.
\newblock Complete sets of reductions for some equational theories.
\newblock {\em J. of the ACM}, 28(2):233--264, 1981.

\bibitem{rao93fsttcs}
K.~Rao.
\newblock Completeness of hierarchical combinations of term rewriting systems.
\newblock In {\em Proc. of FSTTCS'93\em, LNCS 761}.

\bibitem{focal0.3.1}
R.~Rioboo, D.~Doligez, T.~Hardin, and all.
\newblock {\em {FoCal} Reference Manual, version 0.3.1}.
\newblock Universit\'e Paris 6, CNAM \& INRIA, 2005.
\newblock \url{http://focal.inria.fr/}.

\bibitem{thompson86lfp}
S.~Thompson.
\newblock Laws in {M}iranda.
\newblock In {\em Proc. of LFP'86}.

\bibitem{thompson90scp}
S.~Thompson.
\newblock Lawful functions and program verification in {M}iranda.
\newblock {\em Science of Computer Programming}, 13(2-3):181--218, 1990.

\bibitem{wadler87popl}
P.~Wadler.
\newblock Views: a way for pattern matching to cohabit with data abstraction.
\newblock In {\em Proc. of POPL'87}.

\bibitem{weis03}
P.~Weis.
\newblock Private constructors in {OC}aml.
\newblock \url{http://alan.petitepomme.net/cwn/2003.07.01.html#5}, 2003.

\end{thebibliography}
